---
title: "Final Project"
output: html_document
date: '2022-05-10'
---

This section pulls the intital data and separates it into normal and log transformed
```{r}
library(tidyverse)
library(dplyr)
library(leaps)
library(zoo)

data <- read_csv("/Users/laurenmctigue/Desktop/Columbia/Machine\ Learning/Project/Project\ Data/hf315-01-ne-towns-panel.csv")

dim(data)
#rename key columns
data <- data %>% rename(
  PercentProtected = co.pre100pct,
  MedianIncome = mhhi.ia.pre
)

na_vals <- data[is.na(data$MedianIncome),]
no_vals <- as.data.frame(na_vals)
x <- no_vals %>% filter(year == "2015") #only one town where the missing value is 2015, can interpolate

interpolated_data <- data %>% filter(town.id != 507) %>%
  mutate(`MedianIncome` = na.approx(`MedianIncome`))

interpolated_data$gis.join <- as.factor(interpolated_data$gis.join)
interpolated_data$year <- as.factor(interpolated_data$year)

#split data into log and normal
log_data <- interpolated_data %>% dplyr::select(year, gis.join,starts_with("ln."), co.pre100:l5.nn10.co.pre100) 

#rename log columns
log_data <- log_data %>% rename(
  Unemployment_rate = ln.unempr.pre,
  Median_Income = ln.mhhi.ia.pre,
  Ag_Employment = ln.agricu.pre ,
  Arts_Employment = ln.arts.pre,
  Pop = ln.popcen.pre,
  All_Protected = co.pre100,
  Private_Protected = co.priv.pre100,
  Public_Protected = co.pub.pre100,
  Distance_100 = ln.ctydist100,
  Distance_30 = ln.ctydist30
)
```

Exploring missing values, performs interpolation for variable of interested
```{r}
na_vals <- log_data[is.na(log_data$Median_Income),]
no_vals <- as.data.frame(na_vals)
x <- no_vals %>% filter(year == "2015") #none
x1 <- no_vals %>% filter(year == "1990") #none
dim(x1)

na_vals2 <- log_data[is.na(log_data$Unemployment_rate),]
no_vals2 <- as.data.frame(na_vals2)
x2 <- no_vals2 %>% filter(year == "2015") #none
dim(x2)
x3 <- no_vals2 %>% filter(year == "1990") #unemployment is missing from log data
dim(x3)

#Unemployment in original data set
unemployment_test <- as.data.frame(data[is.na(data$unemp.r.pre),])
head(unemployment_test)
x4 <- unemployment_test %>% filter(year == "1990")
dim(x4)
#values missing from 1990, cannot use

interpolated_data <- log_data %>% mutate(`Median_Income` = na.approx(`Median_Income`))

interpolated_data$gis.join <- as.factor(interpolated_data$gis.join)
interpolated_data$year <- as.factor(interpolated_data$year)
```


Going to use 1995 and on since it allows more data points
Allows more variables and better results

```{r}
dim(log_data)
#remove 1990
log_data_trim1 <- log_data %>% mutate(`Median_Income` = na.approx(`Median_Income`))
log_data_trim <- log_data_trim1 %>% dplyr::filter(year!=1990)
sapply(log_data_trim, function(x) sum(is.na(x)))

#remove values that are empty still leaves a good amount of data
log_data_trim_selected <- log_data_trim %>% dplyr::select(-ln.r.units.pre, -Ag_Employment, -Arts_Employment, -Pop)
dim(log_data_trim_selected)

#split into test and train data  
train = sample(1:nrow(log_data_trim_selected), nrow(log_data_trim_selected)/2) #split in half
data.train=log_data_trim_selected[train,]
data.test=log_data_trim_selected[-train,]
dim(data.test)
dim(data.train) #3750 in each

simple_regression <- lm(Median_Income ~ All_Protected, data = data.train)
summary(simple_regression)

multi_regression <- lm(Median_Income ~ . - year -gis.join, data = data.train)
summary(multi_regression)

#only use sig p values
multi_regression1 <- lm(Median_Income ~ Unemployment_rate + Distance_100 + ln.pop.cen90 +
                          l5.co.pre100 + l5.co.pub.pre100 + l5.co.priv.pre100, data = data.train)
summary(multi_regression1)

multi_regression2 <- lm(Median_Income ~ Unemployment_rate + Distance_100 + 
                          l5.co.pre100 + l5.co.pub.pre100 + l5.co.priv.pre100, data = data.train)
summary(multi_regression2)

multi_regression3 <- lm(Median_Income ~ Unemployment_rate + Distance_100 + l5.co.pub.pre100 + l5.co.priv.pre100, data = data.train)
summary(multi_regression3)
#all protected not included

#best subset
regfit.full=regsubsets(Median_Income ~ . - year -gis.join,data=data.train,nvmax=19)
reg.summary=summary(regfit.full)
names(reg.summary)
reg.summary$rsq
which.max(reg.summary$adjr2)
plot.new()
par(mfrow=c(2,2))
points(10,reg.summary$adjr2[10], col="red",cex=2,pch=20)
plot(reg.summary$cp,xlab="Number of Variables",ylab="Cp",type='l')
which.min(reg.summary$cp)
points(10,reg.summary$cp[10],col="red",cex=2,pch=20)
which.min(reg.summary$bic)
plot(reg.summary$bic,xlab="Number of Variables",ylab="BIC",type='l')
points(9,reg.summary$bic[9],col="red",cex=2,pch=20)
plot(regfit.full,scale="r2")
plot(regfit.full,scale="adjr2")
plot(regfit.full,scale="Cp")
plot(regfit.full,scale="bic")
coef(regfit.full,10)
coef(regfit.full,9)

#forward selection
regfit.fwd=regsubsets(Median_Income ~ . - year -gis.join,data=data.train,nvmax=19,method="forward")
summary(regfit.fwd)
#distance 100 always included, then private and then public protected

#backward selection
regfit.bwd=regsubsets(Median_Income ~ . - year -gis.join,data=data.train,nvmax=19,method="backward")
summary(regfit.bwd)
#similar to forward

#important variables and meaning
#l5.co.pre100 = 5 year lag of log percent protected, all types
#l5.co.pub.pre100 = 5 year lag of log percent protected, public
#l5.co.priv.pre100 = 5 year lag of log percent protected, private

#try linear of 5 yr lag
simple_regression1 <- lm(Median_Income ~ l5.co.pre100, data = data.train)
summary(simple_regression1)

simple_regression2 <- lm(Median_Income ~ l5.co.pub.pre100 + l5.co.priv.pre100 , data = data.train)
summary(simple_regression2)

```


Test to try year as a cateogrical variable- nothing too signficiant but wanted to see the relationship
```{r}

#year as categorical
data.train.categorical <- data.train
data.train.categorical$year.f <- factor(data.train.categorical$year)
is.factor(data.train.categorical$year.f)

model.cat1 <- lm(Median_Income ~ year.f, data = data.train.categorical)
summary(model.cat1)
#shows a sig p value at 2005 and 1995
#all slopes positive

model.cat2 <- lm(Median_Income ~ year.f + l5.co.pub.pre100, data = data.train.categorical)
summary(model.cat2)

model.cat3 <- lm(Median_Income ~ year.f + l5.co.pub.pre100 + l5.co.priv.pre100, data = data.train.categorical)
summary(model.cat3)
#signs start to flip for the year

```

Year as numerical - nothing too significant

```{r}
#year as numerical
model.num1 <- lm(Median_Income ~ year, data = data.train)
summary(model.num1)

model.num2 <- lm(Median_Income ~ year + l5.co.pub.pre100, data = data.train)
summary(model.num2)
model.num3 <- lm(Median_Income ~ year + l5.co.pub.pre100+ l5.co.priv.pre100, data = data.train)
summary(model.num3)

```

